Slavnov products and KP tau functions

### 2025-04-28 draft.tex
improve draft

\documentclass[a4paper,11pt]{amsart}

\usepackage{/home/thiago/.config/dot-files/latex/globaldef}
\input{/home/thiago/.config/dot-files/latex/definitions.tex}

@@

The study of connections between classical and quantum integrable
systems is an old subject, and has been addressed from a wide range of
perspectives~\cite{Wu:1975mw, Its:1992bj, korepin2000the, Foda2009,
  Alexandrov:2011aa}.  In~\cite{Foda:2009zz}, the authors started the
analysis of the connections between the Slavnov products of the
Heisenberg XXZ spin chain, and the Kadomtsev–Petviashvili (KP)
classical integrable hierarchy. This analysis has been extended to
some directions and applications, for example~\cite{Wheeler:2010vmq,
  Foda:2010, Takasaki:2010qm, Foda:2012wn, Foda:2012wf}.

In (my papers)
the author of the current paper started studying some other cases of
this interesting connection, in particular, in the case of the Temperley-Lieb spin chains
and the Q-Boson model. In all these cases, although the specific details of he
Slavnov product are radically different, the general functional structure
are basically the same, modulo some multiplicative terms.

In~\cite{Belliard:2019bfz}, the authors show
the reason behind this fact.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Scalar products and determinant formulas}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section reviews the main arguments in~\cite{Belliard:2019bfz}.

Let us start with a set of arbitrary complex numbers
\(\bm{u} = \{u_j\}_{j=1}^{n+1}\), and we define \(n+1\) sets
\(\bm{u}_j=\bm{u}\setminus \{u_j\}\) so that we can define
\(n+1\) off-shell Bethe vectors \(\ket{\Psi(\bm{u}_j)}\).

Consider now a set of Bethe roots \(\bm{v} = \{ v_k \}_{k=1}^n\),
so that we have the on-shell Bethe vectors \(\ket{\Psi(\bm{v})}\). 
In fact, let us consider the action of the transfer matrix \(\mathcal{T}(z)\)
on the dual Bethe vector, that is 
\begin{equation}
\label{eq:eigenvalue-bethe}
\bra{\Psi(\bm{v})}\mathcal{T}(z) = \Lambda(z; \bm{v}) \bra{\Psi(\bm{v})} \; , 
\end{equation}
where \(\Lambda(z; \bm{v}) \) is the eigenvalue of the transfer matrix.

From these definitions, we now introduce \(n+1\) Slavnov products 
\begin{equation}
  \bm{\zeta}_j = \bracket{\Psi(\bm{v})}{\Psi(\bm{u}_j)}\qquad j =1, \dots, n+1\; .
\end{equation}
We want to show that \(\bm{\zeta}_j\) satisfy a system of linear equations.

In order to see this, we calculate
\begin{equation}
  \bra{\Psi(\bm{v})} {\color{violet}\left(\mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)} \right)} =
  {\color{violet}\left(\bra{\Psi(\bm{v})}\mathcal{T}(u_j)\right)}\ket{\Psi(\bm{u}_j)} 
  \qquad j =1, \dots, n+1\; .
\end{equation}
The right-hand side of this equation can be simplified with the eigenvalue
expression~(\ref{eq:eigenvalue-bethe}), that is
\begin{equation}
  {\color{violet}\left(\bra{\Psi(\bm{v})}\mathcal{T}(u_j)\right)}\ket{\Psi(\bm{u}_j)}  =
\Lambda(u_j; \bm{v}) \bracket{\Psi(\bm{v})}{\Psi(\bm{u}_j)} = 
\Lambda(u_j; \bm{v}) \bm{\zeta}_j\; . 
\end{equation}

In order to simmplify the left-hand side, we expand the action of the transfer matrix on
the off-shell as
\todo{\tiny This point seems to be the most delicate. It does not work for
  elliptic cases}
\begin{equation}
  \mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)}  = \sum_{k=1}^{n+1} L_{jk} \ket{\Psi(\bm{u}_k)} \; ,
\end{equation}
where \(L_{jk}\) are numerical coefficients. Observe that the terms \(L_{jk}\) with
\(j\neq k\) contain the unwanted terms. In the particular case where the algebraic
Bethe equations are satisfied, these unwanted terms vanish and only the diagonal
coefficient survives, that is \(L_{jj}\), and it is equal to the eigenvalue of the
tranfer matrix.

Therefore, we have
\begin{equation}
  \bra{\Psi(\bm{v})} {\color{violet}\left(\mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)} \right)} =
  \sum_{k=1}^{n+1} L_{jk} \bm{\zeta}_k \; .
\end{equation}

Puttting all these facts together, we get
\begin{equation}
  \sum_{k=1}^{n+1} L_{jk} \bm{\zeta}_k  = 
  \Lambda(u_j; \bar{v}) \bm{\zeta}_j\quad \Rightarrow \quad
  \sum_{k=1}^{n+1} M_{jk} \bm{\zeta}_k = 0 \; ,
\end{equation}
where
\begin{equation}
  M_{jk} = L_{jk} - \delta_{jk}\Lambda(u_j; \bm{v})\; . 
\end{equation}
Here is the system we were looking for. We can write this system in
a matrix form as 
\begin{equation}
  \left(\bm{L} - \bm{\Lambda} \right) \cdot \bm{\zeta} = \bm{0}\; ,
\end{equation}
where
\begin{subequations}
  \begin{equation}
    \bm{L} = 
 \begin{pmatrix}
   L_{1,1} & \cdots & L_{1,n+1}\\
   L_{2,1} & \cdots & L_{2,n+1}\\
   \vdots & \vdots & \vdots\\
   L_{n+1, 1} & \cdots & L_{n+1, n+1}\\
 \end{pmatrix}  \quad , \quad
    \bm{\Lambda} = \mathrm{diag}(\Lambda_1, \Lambda_2, \dots, \Lambda_{n+1})
\end{equation}
where we have written \( \Lambda_j \equiv \Lambda(u_j; \bm{v})\), and also
\begin{equation}
  \bm{\zeta} = (\bm{\zeta}_1, \dots , \bm{\zeta}_{n+1})^T\qquad \, 
  \bm{0} = (0, \dots , 0)^T\; .
\end{equation}
\end{subequations}

The whole point now is that if this system has a solution,
it will be written as minors of the matrix \(M\).

@@

We will now consider some concrete models. In the original
work~\cite{Belliard:2019bfz}, the authors consider solutions with a
rational R-mnatrix with periodic boundary conditions. Here we want to
review the main results of this work. 
\todo{\tiny Extend these results to the trigonometric case}

We start with an R-matrix of the form
\begin{equation}
  R(u,v) = \mathbb{1} + g(u,v) \mathcal{P}\; , \qquad
  g(u,v) = \frac{c}{u - v}
\end{equation}
where \(c\) is a constant. Moreover \(\mathbb{1}\) is the identity operator and
\(\mathcal{P}\) is the permutation operator.

@@

Let us write the \(p\) functions \(\sigma_p\) in terms of a linearly
\todo{Not sure if I need this!!!}
independent basis \(\{\phi_j\}_{j=1}^n\). In this case, we have that the functions
\(\mathcal{Y}_j(z)\) are also linearly independent and have the expansion
\begin{equation}
  \mathcal{Y}_j(z) = \sum_{k=1}^n \rho_k(\bm{v}_j) \phi_k(z)\; ,
\end{equation}
where the coefficients \( \rho_k(\bm{v}_j)\) depend on the Bethe roots \(\bm{v}_j\). 
The functional structure of the coefficients \(\rho_k\), when written, in terms of the
Bethe roots are the same.

For example, in the case of the XXX model, \(\alpha_p(z)\sim z^{2n} + \cdots\), so
that one can take \(\phi_j \sim z^{2j} + \cdots\), and the coefficients depend on the
Bethe roots and inhomogeneities.

@@

It seems that we can consider a basis of functions \(\mathcal{Y}_j(z)\) for the Baker-Akhiezer
functions. I think that this object might say something interesting. In particular, we can
explicitly build solutions of the KP equation. In fact, the functions \(\mathcal{Y}_j\)
above have poles at the Bethe roots, it is interesting. 
\begin{itemize}
\item[\faCheck] What is the pole and root structure of the tau and Baker-Akhiezer (BA) functions?
  In particular, what happens at the Bethe roots?
\item[\faCheck] Basis tau functions
\item[\faCheck] Write the cases n=2 and n=3
\item[\faCheck] What the general case for all \(z_k\) approaching to a single
  value \(w\). 
\item[\faCheck] Write these tau functions in terms of Miwa coordinates
\item[\faCheck] I need to calculate the Baker-Akhiezer function for this model using the Japanese formula
  in terms of tau functions.
\item[\faCheck]  Write some explicit expressions for \(N=2\) and \(N=3\). 
\item See the Limit \(L \to \infty\) and \(N \to \infty\) but sufficiently slower than \(L\). 
  I think that the matrix \(\mathcal{T}\) is a trace class operator and we write the BA function
  as a Fredholm determinant. 
\item Better investigate these functions in Miwa coordinates. See Fredholm determinant. 
  I think that this BA function will be similar to the soliton-like solution discussed in section 3.6.2 (page 41) of~\cite{Zabrodin2018}. 
  My functions seem to have a pole of order \(N\) at infinity. And \(N\) zeros at the Bethe roots. 
\item It turns out that the Baker-Akhiezer function is the appropriate object to discuss
  the Bethe roots. 
  \item I think that these tau functions come from algebro geometric tau functions. 
  \item I can try to study the dynamics closer to the Bethe roots, I think that the
    \(\Omega\)'s will have some interesting expansions, some of them might vanish. 
  \item This might become a Riemann Hilbert problem
  \item Are these tau functions associated to the Calogero-Moser model??
  \item Elliptic case - see slavnov, zabrodin's paper
\end{itemize}

>>>>

\documentclass[a4paper,12pt]{amsart}

\usepackage{globaldef}
\input{definitions.tex}

@@

The study of connections between classical and quantum integrable
systems is an old subject, and has been addressed from a wide range of
perspectives~\cite{Wu:1975mw, Its:1992bj, korepin2000the, Foda2009,
  Alexandrov:2011aa}.  In~\cite{Foda:2009zz}, the authors started the
analysis of the connections between the Slavnov products of the
Heisenberg XXZ spin chain, and the Kadomtsev–Petviashvili (KP)
classical integrable hierarchy. This analysis has been extended to
some directions and applications, for example~\cite{Wheeler:2010vmq,
  Foda:2010, Takasaki:2010qm, Foda:2012wn, Foda:2012wf}.
In~\cite{Araujo:2021ghu, Araujo:2024klz}, the author of the current
paper also investigated some developments of this research field. In
particular, the relation between integrable hierarchies and quantum
integrable systems such as the Q-Boson integrable system and in the
Temperley-Lieb open spin chains.

Although the specific details of he Slavnov product are radically
different, In all these cases discussed in the previous paragraph, the
general functional structure of the Slavnov products of different
models are basically the same. More specifically, module some
multiplicative terms, all these expressions are determinant, and it is
the fact that allows us to manipulate these objects to show that thesy
are KP and Toda tau functions.

Belliard and Slavnov~\cite{Belliard:2019bfz} have addressed precisely
this property: \emph{Why scalar products in the algebraic Bethe ansatz
have determinant representations}.  Using their results, we now want
to understand \emph{why and how (some) scalar products in the
algebraic Bethe ansatz are KP and Toda tau functions}. In the current
show that as long as the integrable system satisfies the conditions
established in~\cite{Belliard:2019bfz} -- and some additional
restrictions -- the Slavnov product is guaranteed to satisfy the KP
integrable hiearchy. Moreover, we also discuss some immediate
consequences of this result.

In section 2 we review the main aspects of the
work~\cite{Belliard:2019bfz} and fix the main notation we use in the
remainder of this work. (...)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Scalar products as determinants}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In order to make the paper as self-contained as possible, this section
reviews the main arguments in~\cite{Belliard:2019bfz}. Let us start
with a set of arbitrary complex numbers \(\bm{u} =
\{u_j\}_{j=1}^{n+1}\), and we set \(n+1\) sets
\(\bm{u}_j=\bm{u}\setminus \{u_j\}\). In the algebraic Bethe ansatz
setting~\cite{Korepin:1993kvr}, we can now define \(n+1\) off-shell
Bethe vectors \(\ket{\Psi(\bm{u}_j)}\), i.e. Bethe states where the
algebraic Bethe equations are not imposed on
\(\bm{u}_j\). Additionally, we also consider a set of Bethe roots
\(\bm{v} = \{ v_k \}_{k=1}^n\), and the on-shell Bethe vectors
\(\ket{\Psi(\bm{v})}\).

\subsection{Determinant representation}
The action of the transfer matrix \(\mathcal{T}(z)\), that is an
Hermitian operator, on the dual Bethe vector is given by
\begin{equation}
\label{eq:eigenvalue-bethe}
\bra{\Psi(\bm{v})}\mathcal{T}(z) = \Lambda(z; \bm{v}) \bra{\Psi(\bm{v})} \; , 
\end{equation}
where \(\Lambda(z; \bm{v}) \) is the eigenvalue of the transfer matrix.

We now introduce \(n+1\) scalar products between the on-shell and
off-shell Bethe states
\begin{equation}
  \bm{\zeta}_j(\bm{u}_j, \bm{v}) \equiv  \bm{\zeta}_j
  = \bracket{\Psi(\bm{v})}{\Psi(\bm{u}_j)}\qquad j =1, \dots, n+1\; .
\end{equation}
These partially on-shell scalar products are called \emph{Slavnov
products}.

Before procceding, let us also establish the class of models we are
dealing with.  Let us consider a class of integrable models where the
action of the transfer matrix on a generic off-shell bethe states can
be expanded as
\begin{subequations}
\begin{equation}
\label{eq:scalar-exp}
  \mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)}  = \sum_{k=1}^{n+1} L_{jk} \ket{\Psi(\bm{u}_k)} \; ,
\end{equation}
where \(L_{jk}\) are numerical coefficients. Observe that the terms
\(L_{jk}\) with \(j\neq k\) contain the unwanted terms.; and these
non-diagonal terms must vanish when the Bethe equations are
satisfied. Therefore, since the diagonal coefficient survives, that is
\(L_{jj}\), we conclude that these terms must be equal to the
eigenvalue of the tranfer matrix, that is
\begin{equation}
  \mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)} =  \Lambda(u_j; \bm{u}_j) \ket{\Psi(\bm{v})} + 
  \sum_{\substack{k=1\\ k\neq j}}^{n+1} L_{jk} \ket{\Psi(\bm{u}_k)} \; . 
\end{equation}
\end{subequations}
This class of models include a wide range of known integrable systems,
including the ones discussed in the introduction. One important class
of models that does not fall in this classification are with elliptic
R-matrices, although some of these models have slavnov products with
determinantal representations.

\begin{proposition}
The Slavnov products \(\bm{\zeta}_j\) satisfy a system of linear
equations.
\end{proposition}

\paragraph{\emph{Proof}} Let us first use that
\begin{equation}
\label{eq:scalar-prod}
  \bra{\Psi(\bm{v})} \left(\mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)} \right) =
  \left(\bra{\Psi(\bm{v})}\mathcal{T}(u_j)\right)\ket{\Psi(\bm{u}_j)}
  \qquad j =1, \dots, n+1\; .
\end{equation}
The right-hand side of this equation can be simplified with the eigenvalue
expression~(\ref{eq:eigenvalue-bethe}), that is
\begin{equation}
  \left(\bra{\Psi(\bm{v})}\mathcal{T}(u_j)\right)\ket{\Psi(\bm{u}_j)}  =
\Lambda(u_j; \bm{v}) \bracket{\Psi(\bm{v})}{\Psi(\bm{u}_j)} = 
\Lambda(u_j; \bm{v}) \bm{\zeta}_j\; . 
\end{equation}

Let us now use the expansion~(\ref{eq:scalar-exp}) on the left-hand
side of~(\ref{eq:scalar-prod}). Therefore, we have
\begin{equation}
  \bra{\Psi(\bm{v})} \left(\mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)} \right) =
  \sum_{k=1}^{n+1} L_{jk} \bm{\zeta}_k \; .
\end{equation}

Puttting all these facts together, we get
\begin{equation}
  \sum_{k=1}^{n+1} L_{jk} \bm{\zeta}_k  = 
  \Lambda(u_j; \bar{v}) \bm{\zeta}_j\quad \Rightarrow \quad
  \sum_{k=1}^{n+1} M_{jk} \bm{\zeta}_k = 0 \; ,
\end{equation}
where
\begin{equation}
  M_{jk} = L_{jk} - \delta_{jk}\Lambda(u_j; \bm{v})\; . 
\end{equation}
And this expression shows that the Slavnov products satisfy a linear system.
We can write this system in a matrix form as 
\begin{equation}
\bm{M} \bm{\zeta} = 
  \left(\bm{L} - \bm{\Lambda} \right) \bm{\zeta} = \bm{0}\; ,
\end{equation}
where
\begin{subequations}
  \begin{equation}
    \bm{L} = 
 \begin{pmatrix}
   L_{1,1} & \cdots & L_{1,n+1}\\
   L_{2,1} & \cdots & L_{2,n+1}\\
   \vdots & \vdots & \vdots\\
   L_{n+1, 1} & \cdots & L_{n+1, n+1}\\
 \end{pmatrix}  \quad , \quad
    \bm{\Lambda} = \mathrm{diag}(\Lambda_1, \Lambda_2, \dots, \Lambda_{n+1})
\end{equation}
with \(\Lambda_j \equiv \Lambda(u_j; \bm{v})\), and also
\begin{equation}
  \bm{\zeta} = (\bm{\zeta}_1, \dots , \bm{\zeta}_{n+1})^T\qquad \, 
  \bm{0} = (0, \dots , 0)^T\; .
\end{equation}
\end{subequations}
 \qed

There are two important consequences for us now. First of all, if we
assume that this system has a non-trivial solution, it will be written
as minors of the matrix \(\bm{M}\).

Moreover, since the system is homogeneous, the solutions are
determined up to multiplicative terms, that we can fix them by
requiring that the final result gives the required tau function. In
the original works of Foda and collaborators, e.g.~\cite{Foda:2009zz,
  Wheeler:2010vmq}, the authors perform a series of redefinitions to
reach the same goal.

Evidently being a determinant is not enough to guarantee that the
Slavnov product is a tau function.  In order to accomplish this
result, we need to specify the models. Following the original
work~\cite{Belliard:2019bfz}, we will only discuss the rational models
-- we will see that the calculations are massive even in this simples
case. The generalization for open boundary conditions and for
trigonometric models is immediate, see~\cite{Slavnov:2019aba}
for more details on the generalization.

@@

We will now consider some concrete models.
We start with an R-matrix of the form
\begin{equation}
  R(u,v) = \mathbb{1} + g(u,v) \mathcal{P}\; , \qquad
  g(u,v) = \frac{c}{u - v}
\end{equation}
where \(c\) is a constant. Moreover \(\mathbb{1}\) is the identity operator and
\(\mathcal{P}\) is the permutation operator.

@@

@@

It seems that we can consider a basis of functions \(\mathcal{Y}_j(z)\) for the Baker-Akhiezer
functions. I think that this object might say something interesting. In particular, we can
explicitly build solutions of the KP equation. In fact, the functions \(\mathcal{Y}_j\)
above have poles at the Bethe roots, it is interesting. 
\begin{itemize}
\item[\faCheck] What is the pole and root structure of the tau and Baker-Akhiezer (BA) functions?
  In particular, what happens at the Bethe roots?
\item[\faCheck] Basis tau functions
\item[\faCheck] Write the cases n=2 and n=3
\item[\faCheck] What the general case for all \(z_k\) approaching to a single
  value \(w\). 
\item[\faCheck] Write these tau functions in terms of Miwa coordinates
\item[\faCheck] I need to calculate the Baker-Akhiezer function for this model using the Japanese formula
  in terms of tau functions.
\item[\faCheck]  Write some explicit expressions for \(N=2\) and \(N=3\). 
\item Discuss the Toda hierarchy case. 
\item See the Limit \(L \to \infty\) and \(N \to \infty\) but sufficiently slower than \(L\). 
  I think that the matrix \(\mathcal{T}\) is a trace class operator and we write the BA function
  as a Fredholm determinant. 
\item Better investigate these functions in Miwa coordinates. See Fredholm determinant. 
  I think that this BA function will be similar to the soliton-like solution discussed in section 3.6.2 (page 41) of~\cite{Zabrodin2018}. 
  My functions seem to have a pole of order \(N\) at infinity. And \(N\) zeros at the Bethe roots. 
\item It turns out that the Baker-Akhiezer function is the appropriate object to discuss
  the Bethe roots. 
  \item I think that these tau functions come from algebro geometric tau functions. 
  \item I can try to study the dynamics closer to the Bethe roots, I think that the
    \(\Omega\)'s will have some interesting expansions, some of them might vanish. 
  \item This might become a Riemann Hilbert problem
  \item Are these tau functions associated to the Calogero-Moser model??
  \item Elliptic case - see slavnov, zabrodin's paper
\end{itemize}
